---
title: "Learning Advanced Row Grouping in the tidyverse: learning rowwise() and nest-by()"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
penguins <- readRDS("data/penguins.rds")
```


You’re probably familiar with `group_by()` as a way to calculate summaries over a set of groups in your data. 

However, there are many other techniques in the tidyverse for grouping data that are extremely helpful when doing things like modeling, or resampling data. 

Here’s a common workflow for doing things with `rowwise()`:

1. Get your data into a list-column format with `nest_by()` or `resample()`
2. Pipe your data into `rowwise()`
3. Pipe `rowwise()` into `summary()`, `mutate()`, `do()`, or `map()` to do something with each row of the data.

### Reminder: Penguins Data


```{r}
penguins
```

## `rowwise()` - running from a metadata frame about samples

`rowwise()` lets you run operations one row at a time on the data. It’s a lot like `group_by()`, but it groups the data.frame by row, rather than group.

This is especially useful when you have wide data where multiple columns need to be summarized. 

### `rowwise() %>% summarize()`

```{r summ, exercise=TRUE}
penguins %>%
  rowwise() %>%
  summarize(species, island, 
            average_bill = mean(c(bill_length_mm, bill_depth_mm)))
```

### Try it Out

Calculate the `rowwise()` sum of `bill_length_mm` and `flipper_length_mm` as `sum_mm`. 

```{r rowsum, exercise=TRUE}
penguins %>%
  rowwise() %>%
  summarize(species, island,
            sum_mm = ---------)
```

### `rowwise() %>% mutate()`

In a `rowwise` workflow, you often want to apply a function to multiple columns. 

```{r mutate, exercise=TRUE}
penguins %>%
  rowwise() %>%
  select(species, island, bill_length_mm, bill_depth_mm) %>%
  mutate(max_bill = max(
    c(bill_length_mm, bill_depth_mm)
    ))

```




## `rowwise()` and `tidyselect`: `c_across()`

You might think that you can use `across()` to select columns with `tidyselect` helpers such as `starts_with()` in a `rowwise()` workflow. 

Well, almost. You need to sub in `c_across()` for `across()`. Here we're applying `min()` to find the minimum value of any column that contains "bill":

```{r cacross, exercise=TRUE}
penguins %>%
  rowwise() %>%
  select(species, island, bill_length_mm, bill_depth_mm) %>%
  mutate(min_bill = min(c_across(contains("bill"))))
```


### Try it Out

```{r}

```


## `nest_by()` 

Let's talk about restructuring our data to fit into a `rowwise()` workflow. One way we can do this is by packaging data into list-columns. A list column can contain a subsetted dataset, which makes it easier to do things such as run multiple models on different subsets of data.

`nest_by()` will take your data and put it into a *list-column* - you can think of a list-column as a vector, where each entry is another data frame.

```{r nesting, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species)

nested_penguins
```

Note that the list-column that contains the data is called `data`. You can change this name by specifying the `.key` argument. 

How do we get data out? We can use `[[]]` (double brackets) to extract the data directly from a row:

```{r leaving-nest, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species)

nested_penguins$data[[1]]
```

Data in a list column format can be really easily utilized by a `rowwise()` workflow. Here, we're counting the number of rows in each row of `data`:

```{r row_nest, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species,sex) %>%
  rowwise() %>%
  mutate(n = nrow(data))

nested_penguins
```

### Try it out

Find the maximum `bill_length_mm` within each row. Hint: you may have to use the `$` operator on `data` to access `bill_length_mm` for each dataset. 

```{r max_bill, exercise=TRUE}
penguins %>%
  nest_by(species,sex) %>%
  rowwise() %>%
  summarize(species, sex,
    max_length = max(data$bill_length_mm))

```

## So What?

We've only done things that you can easily do with a `group_by()/summarize()` statement so far. Where `rowwise()/nest_by()` becomes extremely powerful is that you can run multiple models on data in a list-column.

### Running lots of models on list-columns using `rowwise()`

List columns really come into their own when you need to build multiple models with different subsets of the data. 

Because we noted the number of rows in `data` for each row grouping, we can now filter on that before we apply a model. 

We use `broom` here to extract the `p.value` from the model result

```{r model_nest, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species,sex) %>%
  mutate(n = nrow(data))

model_penguins <- nested_penguins %>%
  filter(n > 6) %>%
  mutate(model = list(lm(bill_length_mm ~ flipper_length_mm, data=data))) %>%
  summarize(n, broom::tidy(model)) %>%
  filter(term == "flipper_length_mm")

model_penguins %>%
  filter(p.value < 0.05)
```

### Exercise

Use `nest_by()` to nest by data by `island` and run this linear model:

```
lm(bill_depth_mm ~ bill_length_mm, data=data)
```

```{r nest2, exercise=TRUE}
penguins %>%
  nest_by(------) %>%
  mutate(model = -------------)
```


### Exercise

Run these two different models on the data and call them `model1` and `model2`. Extract the p-values for each using `broom::tidy()`

```
lm(bill_depth_mm ~ bill_length_mm, data=data)
```

and

```
lm(bill_length_mm ~ flipper_length_mm, data=data)
```

```{r}

```


More info about list columns here: https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html

Much more about applying many models to data here: https://r4ds.had.co.nz/many-models.html
