---
title: "Learning rowwise() and nest_by()"
output: learnr::tutorial
runtime: shiny_prerendered
description: Learn about rowwise() and nest_by()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(learnr)
penguins <- readRDS("data/penguins.rds")
gap <- readRDS("data/gap.rds") %>%
  mutate(StudentId = paste0("SID", StudentId))
```

## Learning Objectives

- **Learn** and **apply** `rowwise()` to produce row-by-row summaries
- **Understand** what `c_across()` does
- **Use** `nest_by()` to make list-columns
- **Access** the `data` in a list-column and **do** things with it.
- **Run** multiple statistical models on a list-column using `rowwise()`


You’re probably familiar with `group_by()` as a way to calculate summaries over a set of groups in your data. 
However, there are many other techniques in the tidyverse for grouping data that are extremely helpful when doing things like modeling, or resampling data. 


### Reminder: Penguins Data


```{r}
penguins
```


## Intro to `rowwise()` 

`rowwise()` lets you run operations one row at a time on the data. It’s a lot like `group_by()`, but it groups the data.frame by row, rather than group. This is especially useful when you have wide data where multiple columns need to be utilzed. 

Here’s a common workflow for doing things with `rowwise()`:

1. Get your data into a list-column format with `nest_by()` or `resample()`
2. Pipe your data into `rowwise()`
3. Pipe `rowwise()` into `mutate()` to do something with each row of the data.


### `rowwise() %>% mutate()`

One common pattern is to use `rowwise()` to calculate per-row variables. 

```{r summ, exercise=TRUE}
penguins %>%
  rowwise() %>%
  mutate(average_bill = mean(c(bill_length_mm, bill_depth_mm))) %>%
  select(species, island, average_bill)
```

Compare this to not using `rowwise()`:

```{r summ2, exercise=TRUE}
penguins %>%
  mutate(average_bill = mean(c(bill_length_mm, bill_depth_mm))) %>%
  select(species, island, average_bill)
```

### Exercise

Calculate the `rowwise()` sum of `bill_length_mm` and `flipper_length_mm` as `sum_mm`. 

```{r rowsum, exercise=TRUE}
penguins %>%
  rowwise() %>%
  mutate(sum_mm = ---------) %>%
  select(species, island, sum_mm)
```

```{r rowsum-solution}
penguins %>%
  rowwise() %>%
  mutate(sum_mm = bill_length_mm + flipper_length_mm) %>%
  select(species, island, sum_mm)
```


## `rowwise()` and `tidyselect`: `c_across()`

You might think that you can use `across()` to select columns with `tidyselect` helpers such as `starts_with()` in a `rowwise()` workflow. 

Well, almost. You need to sub in `c_across()` for `across()`. 

Here we're applying `min()` to find the minimum value of any column that contains "bill":

```{r cacross, exercise=TRUE}
penguins %>%
  rowwise() %>%
  select(species, island, bill_length_mm, bill_depth_mm) %>%
  mutate(min_bill = min(c_across(contains("bill"))))
```


### Exercise

Find the minimum value of any of the numeric columns using `where(is.numeric)`:

```{r min1, exercise = TRUE}
penguins %>%
  rowwise() %>%
  mutate(min_val = min(---------))
```

```{r min1-solution}
penguins %>%
  rowwise() %>%
  mutate(min_val = min(c_across(where(is.numeric))))
```


### `rowwise()` can be useful with wide data

Sometimes you really don't want to deal with making a long data frame from a wide data frame. Usually, your collaborators expect wide data, and you can skip a bit of conversion by utilizing rowwise This is where `rowwise()` can really shine.

Here's the `gap` dataset. One thing we might want to do is calculate the average score for all the numeric variables.

```{r}
head(gap)
```

Here we're using the wide format of this dataset to our advantage.

```{r gap_mean, exercise=TRUE}
gap %>%
  rowwise() %>%
  mutate(mean_score = 
           mean(c_across(where(is.numeric)))
         )
```

## `nest_by()` - making list-columns

Let's talk about restructuring our data to fit into a `rowwise()` workflow. 

One way we can do this is by packaging data into *list-columns*. A cell of a *list-column* can contain a subsetted dataset, which makes it easier to do things such as run multiple models on different subsets of data.

`nest_by()` will take your data and put it into a *list-column* - you can think of a list-column as a vector, where each entry is another data frame.

```{r nesting, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species)

nested_penguins
```

Note that the list-column that contains the data is named `data`. You can change this name by specifying the `.key` argument in `nest_by()`.

### Getting data out of a list-column

How do we get data out? We can use `[[]]` (double brackets) to extract the data directly from a row:

```{r leaving-nest, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species)

nested_penguins$data[[2]]
```

There is also `tidyr::unnest()` to unnest the data completely. You will need to provide a `cols` argument to specify that you want to unnest the `data` column, or R will throw a warning.

```{r leaving-nest2, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species)

nested_penguins %>%
  tidyr::unnest(cols = c(data))
```


### Using list-columns with `rowwise()`

Data in a list column format can be really easily utilized by a `rowwise()` workflow. We've packaged the other columns in a list-column called `data`, so our `mutate()` statements can use `data` to calculate anything we want. 

Here, we're counting the number of rows in each row of `data`:

```{r row_nest, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species,sex) %>%
  rowwise() %>%
  mutate(n = nrow(data))

nested_penguins
```

### Exercise

Use `nest_by()` to nest `penguins` by `species` and `sex`. Find the maximum `bill_length_mm` within each row. Hint: you may have to use the `$` operator on `data` to access `bill_length_mm` for each dataset in the `data` list-column. 

```{r max_bill, exercise=TRUE}
penguins %>%
  nest_by(species,sex) %>%
  rowwise() %>%
  summarize(species, sex,
    max_length = max(---------))

```

```{r max_bill-solution}
penguins %>%
  nest_by(species,sex) %>%
  rowwise() %>%
  summarize(species, sex,
    max_length = max(data$bill_length_mm))

```

## Doing things with list-columns and `rowwise()`

So far, we've only done things that you can easily do with a `group_by()/summarize()` statement so far. 

Where `rowwise()/nest_by()` becomes extremely powerful is that you can run multiple models on data in a list-column.

### Running lots of models on list-columns using `rowwise()`

List columns really come into their own when you need to build multiple models with different subsets of the data. 

Here, we're using `mutate` to run `lm()` on each row of `data`. 

Note that we have to wrap the output of `lm()` with `list()` to get it to "fit" into the list column.

Because we noted the number of rows in `data` for each row grouping, we can now filter on that before we apply a model. 

However, if we access the `model` column, you'll notice that it just `print()`s the output of each fit.

```{r model_nest1, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species,sex) %>%
  mutate(n = nrow(data))

model_penguins <- nested_penguins %>%
  filter(n > 6) %>%
  mutate(model = list(lm(bill_length_mm ~ flipper_length_mm, data=data))) 

model_penguins$model
```


Well, that's a little difficult to utilize. We use `broom::tidy` below to extract the `p.value` from the model result.

```{r model_nesty, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species,sex) %>%
  mutate(n = nrow(data))

model_penguins <- nested_penguins %>%
  filter(n > 6) %>%
  mutate(model = list(lm(bill_length_mm ~ flipper_length_mm, data=data))) %>%
  summarize(species, sex, n, broom::tidy(model)) 

model_penguins
```

Now that we have our tidied results, we can filter. Because `broom::tidy()` has two terms (`flipper_length_mm` and the intercept) for our model, we'll need to filter for our term of interest, `flipper_length_mm`. We can also filter on `p.value`.

```{r model_nest3, exercise=TRUE}
nested_penguins <- penguins %>%
  nest_by(species,sex) %>%
  mutate(n = nrow(data))

model_penguins <- nested_penguins %>%
  filter(n > 6) %>%
  mutate(model = list(lm(bill_length_mm ~ flipper_length_mm, data=data))) %>%
  summarize(species, sex, n, broom::tidy(model)) 
  
model_penguins %>%
  filter(p.value < 0.05) %>%
  filter(term == "flipper_length_mm")
```


### Exercise

Use `nest_by()` to nest by data by `island` and run this linear model on each row of `data`.

```
lm(bill_depth_mm ~ bill_length_mm, data=data)
```

```{r nest2, exercise=TRUE}
penguins %>%
  nest_by(------) %>%
  mutate(model = -------------)
```

```{r nest2-solution}
penguins %>%
  nest_by(island) %>%
  mutate(model = list(lm(bill_depth_mm ~ bill_length_mm, data=data)))
```


### Exercise (this one is tricky!)

Nest the `penguins` data by `island`. Run the two different models below on the data and call them `model1` and `model2`. Extract the p-values for each model using `broom::tidy()`.

`model1`: `lm(bill_depth_mm ~ bill_length_mm, data=data)`

and

`model2`: `lm(bill_length_mm ~ flipper_length_mm, data=data)`

Here's a reminder of what `broom::tidy()` looks for an `lm()` model: 

```{r, echo=TRUE}
tidy_result <- broom::tidy(lm(bill_length_mm ~ flipper_length_mm, data=penguins))

tidy_result
```

Uh-oh. We want to be able to access the `p.value` for the second row. We can do this by using the `$` accessor to access the column, and then request the second value of the column.

In other words, we'll need: 

`tidy_result$p.value[2]`

```{r}
tidy_result$p.value[2]
```

```{r twomodels, exercise=TRUE}
penguins %>%
  nest_by(island) %>%
  mutate(model1 = list(-------),
         model2 = list(-------)
         ) %>%
  mutate(mod1pvalue = broom::tidy(model1)$p.value[2],
         mod2pvalue = ------)
```

```{r twomodels-solution}
penguins %>%
  nest_by(island) %>%
  mutate(model1 = list(lm(bill_depth_mm ~ bill_length_mm, data=data)),
         model2 = list(lm(bill_length_mm ~ flipper_length_mm, data=data))
         ) %>%
  mutate(mod1pvalue = broom::tidy(model1)$p.value[2],
         mod2pvalue = broom::tidy(model2)$p.value[2]
  )
```

## Going Further

More info about list columns here: https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html

Much more about applying many models to data here: https://r4ds.had.co.nz/many-models.html

Much more about row-oriented workflows here: https://github.com/jennybc/row-oriented-workflows
