---
title: "Learning `tidyselect`"
output: learnr::tutorial
runtime: shiny_prerendered
description: Learn more advanced techniques to select columns
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(tidyr)
gap <- readRDS("data/gap.rds")
penguins <- readRDS("data/penguins.rds")
```


## Doing More with Multiple Columns using `tidyselect`

You probably are familiar with doing operations on *rows* (such as `filter`) and on single *columns* (using `mutate`). 

However, a very common use pattern is doing something with *multiple* columns in the data. And the `tidyselect` verbs help you with this. 

Let's look at a couple common patterns for using `tidyselect` in your `dplyr` recipes.

### Learning Objectives

- **Understand** how `tidyselect` helper functions (`contains()`, `all_of()`) work with `select()`.
- **Understand** how `where()` works with `select` and `.is*` functions
- **Learn** how to combine helpers with `c()` and find the *complement* using `-`
- **Apply** `across()` to work with `mutate()` to mutate multiple columns
- **Apply** `across()` with `summarize()` to summarize multiple columns


## `tidyselect` selection helpers

At the heart of `tidyselect` are the *selection helpers*. These let you select variables in a dataset to do operations on, such as `summarize()` or `mutate`.

- `everything()`
- `starts_with()`
- `ends_with()`
- `contains()`
- `any_of()`
- `where()`

We'll see that `where()` is a special case.

Here's a couple of useful things you can do with this.

### `penguins` overview

Here's the `penguins` data for reference:

```{r}
penguins
```


### `contains()`

Select all variables that contain "bill":

```{r bill, exercise=TRUE}
penguins %>%
    select(contains("bill"))
```

Note that by default, `contains()`, `starts_with()`, `ends_with()`, and `match()` are not case sensitive. To enable this, set `ignore.case = FALSE`.

For example, this won't match anything because `"Bill"` is capitalized: 

```{r bill2, exercise=TRUE}
penguins %>%
    select(contains("Bill", ignore.case=FALSE))
```
`starts_with()`, and `ends_with()` are convenient to match the beginning and ending of strings.

```{r}

```

If you want to use regular expressions, `match()` is your friend:

```{r}

```

### `all_of()`

`all_of()` is a strict selector. This statement won't work because `blah` is not a column name in `penguins`:

```{r allof, exercise=TRUE}
selected_variables <- c("island", "year", "species", "blah")

penguins %>%
  select(all_of(selected_variables))
```

This will work:

```{r allof2, exercise=TRUE}
selected_variables <- c("island", "year", "species")

penguins %>%
  select(all_of(selected_variables))
```

What can you use `all_of()` for? If you have a vector or column names, you can use `all_of()` to select and automate a process for those columns.

### `any_of()`

`any_of()` is more permissive that `all_of()`. I have a list of variables in a `character` vector and I want to select them. I can use `any_of()`:

```{r any_of, exercise=TRUE}
selected_variables <- c("island", "year", "species", "blah")

penguins %>%
  select(any_of(selected_variables))
```

### `where()`: a wrapper for `is.*` functions

If you want to use a function like `is.integer` to select columns, you'll need to wrap it in a `where()`. 

```{r where1, exercise=TRUE}
penguins %>%
  select(where(is.integer))
```

### Exercise

Select the columns in `penguins` that are `factor`s.

```{r where2, exercise=TRUE}
penguins %>%
  select(where(--------))
```

```{r where2-solution}
penguins %>%
  select(where(is.factor))
```


### Another use for `where()` 

`where()` can be used with any logical function that can return a `TRUE/FALSE` for an entire vector. 

Note we have to test whether the column is numeric before we can use `mean(.x) > 20` as a criteria, because this expression doesn't work for non-`numeric` vectors.

```{r both, exercise=TRUE}
penguins %>%
  tidyr::drop_na() %>%
  select(where(~ is.numeric(.x) && mean(.x) > 20))
```

## More Useful things to do with `tidyselect` helpers

### Combining Multiple Helpers

You can use `c()` to combine multiple helpers:

```{r cc, exercise=TRUE}
penguins %>%
  select(c(species, contains("bill")))
```

### Using `-` with helpers

You can use the `-` operator to return everything but the helper:

```{r cc2, exercise=TRUE}
penguins %>%
  select(-contains("bill"))
```


## Use `mutate(across())` to do something to multiple columns

`across()` is a function in `dplyr` 1.0 that lets you specify a set of columns that uses a set of criteria. If there is one pattern that makes your life easier, it is `across`.

### Convert all character vectors to factors in a dataframe

The basic syntax:

```
mutate(
  across(
    where(is.character), as.factor
  )
)
```


```{r}
penguins %>%
  group_by(species) %>%
  mutate(across(starts_with("bill"), mean, na.rm=TRUE))
```

Applying a function to every variable that is a factor. Here, we're casting every column that is a `factor` to become `character`:

```{r mutaccwhere, exercise=TRUE}
penguins %>%
   mutate(across(where(is.factor), as.character))
```

### Exercise:

Use `mutate(across())` to change everything that is a `integer` in `penguins` to a `double` using `as.double()`. 

```{r mut_across, exercise=TRUE, exercise.lines=5}
penguins %>%
  mutate(across(where(is.character), as.factor))
```

```{r mut_across-solution}
penguins %>%
  mutate(across(where(is.integer), as.double))
```

### Exercise:

Use `mutate(across())` to convert the Likert scale columns to `ordered` factors in the `gap` dataset. Here is a preview of the first few rows of `gap` to help you out.

```{r}
gap[1:10,]
```


```{r gap1, exercise=TRUE}
gap %>%
  mutate(StudentId = as.character(StudentId)) %>%
  mutate(across(----------, --------))

```


```{r gap1-solution}
gap %>%
  mutate(StudentId = as.character(StudentId)) %>%
  mutate(across(where(is.integer), ordered))

```


## `summarize(across())`

Building on what we've learned, we can use `across()` in `summarize()` to summarize multiple variables.

```{r sum1, exercise=TRUE}
penguins %>%
  group_by(species) %>%
  summarize(across(contains("bill"), mean, na.rm=TRUE))
```

This becomes especially useful when combined with `where`:

```{r sum2, exercise=TRUE}
penguins %>%
  group_by(species) %>%
  summarize(across(where(is.double), mean, na.rm=TRUE))
```

## Other verbs that have `cols` arguments

### `tidyr::pivot_longer()`

Another application that you might have not thought of is using `tidyselect` in `pivot_longer`. 

Using `pivot_longer()` with variables that have "bill" in them. 

```{r pivlong, exercise=TRUE}
penguins %>%
  tidyr::pivot_longer(contains("bill"), names_to="metric", values_to="value")
```

Using `pivot_longer()` with variables that are `numeric`:

```{r pivlong2, exercise=TRUE}
penguins %>%
  tidyr::pivot_longer(where(is.numeric), names_to="metric",
               values_to="value")
```
### `relocate()`: move a group of columns around

The `relocate()` function in `dplyr` also takes `tidyselect` helpers. So if you wanted to move the `bill` columns to be first:

```{r relocate, exercise=TRUE}
penguins %>%
  relocate(contains("bill"))
```


### `unite` multiple columns

The `unite()` function allows you to paste together multiple columns into one. Here, we're making a unique id for each penguin.

```{r unite, exercise=TRUE}
penguins %>%
  mutate(row_num=row_number()) %>%
  tidyr::unite(col="unique_id", 
               #make a unique id of the factor combos and the row number
               c(where(is.factor), row_num))
```

## For More Info

I just realized that I inadvertently reproduced this article by Hadley Wickham.

[Column-wise Operations](https://dplyr.tidyverse.org/articles/colwise.html) from tidyverse.org

This is a more technical description of how `tidyselect` works: 

https://tidyselect.r-lib.org/articles/syntax.html

### Acknowledgements

Thanks to the PDX R User group for all of their questions, especially Aaron M, for asking about case sensitivity of `contains()` and the other matching functions.

### Give Us Feedback

Let us know what you think, whether you found it useful, or whether you found any mistakes in the tutorial. Your name and email are optional, but if you include and we use your feedback, we'll add you as a contributor. 

https://ohsu.ca1.qualtrics.com/jfe/form/SV_50wLZCz9m1AMZTf
