---
title: "Learning purrr"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## What is `purrr`?




### Making your own functions.

Maybe we have a plotting function we want to apply to our list-columns. Here we're making multiple ggplots with each combination of species and sex. 

```{r plot_pingu, exercise=TRUE}
p_plot <- function(data, species, sex){
  ggplot(data) + 
    aes(x=flipper_length_mm, 
    y=bill_length_mm) + 
    geom_point() +
    geom_smooth(method="lm") +
    ggtitle(paste(species, sex, "n=", n))
}
  
nested_penguins <- penguins %>%
  nest_by(species,sex) %>%
  mutate(n = nrow(data))

penguin_plots <- nested_penguins %>%
  filter(n > 6) %>%
  mutate(plots = list(p_plot(data, species, sex)))

penguin_plots[["plots"]]

```



## purrr::map() - when your function doesn’t exist

So far, we’ve managed to avoid defining our own functions. 

But there’s a point where defining your own functions becomes really useful. There is a lot of customization that you can do.

The main rule for using purrr::map() is:

> Figure out how to do it once in a function, then run that function on each unit of the data, and return the appropriate unit of information

### `rowwise() %>% map()`

Let's combine `nest_by()`, `rowwise()`, and `map_df()` into a single workflow.

### Step 1. get it into a list column or a list

The great thing about list-columns is that they retain the information that you need to process each row alongside the data itself.

```{r}


penguins %>%
  nest_by(species) %>%
  rowwise() %>%
  map()


```


### Step 2. write a function that works on each row

### Step 3. use rowwise() %>% map() to apply your function


## related: tune() and tune_grid()

## For more info

Rebecca Barter's post on `purrr::map()` is one of the most succinct and well done that I know of. http://www.rebeccabarter.com/blog/2019-08-19_purrr/
