---
title: "Learning Functions"
author: "You"
date: "2/17/2021"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives

- *learn* why writing functions saves you repeating yourself
- *learn* the basic components of a function
- *write* and *use* a simple function
- *learn* and *apply* tidyeval to make tidy functions

## Don't Repeat Yourself (DRY) - use functions!

If you do something in R once and have to do it again, it's probably best to think of how to do it well once, rather than cutting and pasting again.

For example, say that you have multiple files that have the same format and you want to load them all into R, and glue them into a single dataset.

Part of doing this is by using and writing *functions*. We can take code that we might have to write 10 times in a row, and package it in a way that we can use it multiple times.

Let's learn about how we define a function.

In this line, we define our function by giving it a name, `square_number`, and defining its *arguments* using `function()`. Our function only has one argument, `number`.

    square_number <- function(number) {

What about the `{`? It defines the beginning of a *code block* - a code block defines the code that will run when we use the function.

Now, we calculate the square of our number and assign it to `output_number`.

The next line uses `return()` to return a value from our function. Then we close the code block with a `}`.

```{r}
square_number <- function(number) {
  output_number <- number * number 
  return(output_number)
}

square_number(4)
```

## Your Turn

Make a function called `cube_number` that returns the cube of a number. Test it out and make sure it works.

Hint: you can find the cube of a number using `num^3`.

```{r}
cube_number <- function(num){
  output_number <- -----------
  return(output_number)
}
```


# Functions: Checking for Errors with `if()`

What about when someone tries `square_number("two")`? Since "two" is a character, you can't multiply with it. We probably want them to know that this isn't what we are expecting.

```{r}
square_number("two")
```

We can use `if()` as a way to check if the value of `number` is actually numeric. 

`if()` statements check for a condition in the parentheses, and then execute code in the {} (curly brackets) if true.

What does this look like? Remember `case_when()`?

```{r}
square_number_improved <- function(number) {
  if(!is.numeric(number)){ #check for whether number is numeric
    stop("Your input was not a number")
  }

  output_number <- number * number 
  return(output_number)
  
}

square_number_improved(2)

square_number_improved("two")
```



## Do You Need to Check for All Errors?

It depends on who uses your functions. If they are just for you and your lab, as long as everyone understands what goes into the functions, you probably don't need to anticipate all kinds of errors, just a few major ones.

It's when you are releasing your software as a freely available package that you should think about doing extensive testing for errors, especially the functions you anticipate will be used by most people who use your software.



# When should you write a function?

Programmers have an acronym: DRY, which is short for "Don't Repeat Yourself".

If you plan to do something multiple times, then you should probably think of writing a function.


## Practical Example of a Function


```{r}
count_categories <- function(df){
  counts <- df %>%
    count(disease)
  
  return(counts)
}

```

```{r}
count_categories(smoke_complete)

```

This might be useful if you needed to summarize these disease across a large number of files.

One thing you might notice is that `disease` is hard coded. There is a way to make this more flexible using something called `tidyeval`, which we will learn about next time.

# The magic of `{tidyeval}`

```{r}
count_column <- function(df, column = disease){
  
  df %>%
    count( {{column}} )
  
}

count_column(smoke_complete, column = tumor_stage)
```

```{r}


```


# Exercises

## `load_files()`

What does the function `load_files()` do? 

```{r}
load_files <- function(path="data/smoke_1.xlsx"){
  out_frame <- readxl::read_excel(path, sheet = 1)
  out_frame <- janitor::clean_names(out_frame)
  return(out_frame)
}
```

Write up an example here (You can use one of the excel files in the `data` folder).

```{r}

```

## Count Columns

Write a function called `count_columns()` that takes a `data.frame` as input and returns the number of columns in the `data.frame` (you can use `ncol()` to count the number of columns in a data.frame()). If `df` isn't a `data.frame`, then have the function `stop()` and tell the user that they didn't input a `data.frame`.

```{r}
count_columns <- function(df){
  if(!is.data.frame(df)){
    
  }
  
  number_of_cols <- --------------
  return(number_of_cols)
}
```

Show an example of your function in use:

```{r}

```


