---
title: "Learning Lists"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(purrr)
```

## Learning Objectives

- **Learn** about lists as a general purpose data structure
- **Learn** and **utilize** list properties 
- **Access** list elements using `$` and `[[]]`
- **Understand** the difference between homogeneous and heterogeneous lists
- **Use** `purrr::pluck()` to access list elements
- **Understand** how `data.frames()` are list-like

## What are lists?

Lists are a general purpose data structure in R. They are very flexible, in that they have *slots*. A *slot* can contain anything: it could contain a `vector`, a `data.frame`, or even another list!

Unlike vectors, a `list` can also be heterogeneous: each slot can contain a different data type.

We'll start with a making a list and understand different properties about it.

```{r list1, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list
```

`my_list` has some properties and ways to access things in it that are really useful. The first is that `my_list` has a `length()`

```{r list2, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

length(my_list)
```

Additionally, because the elements are named, it also will give you the names:

```{r list3, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

names(my_list)
```

We can access a single element in the list using a name and `$`:

```{r list4, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list$cat_names
```

You can also access an element using the `[[]]` (double brackets) accessor:

```{r list5, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list[[1]]
```

You can also pass a name here as a `character`:

```{r list6, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list[["hedgehog_names"]]
```

Beware of the difference between `[[]]` and `[]`. The `[]` accessor returns a list of length 1 that is named `hedgehog_names`, not the element in it:

```{r list7, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list["hedgehog_names"]
```

In most cases, you want to use `[[]]`.

### Your Turn

Find the length of the `cat_names` element in `my_list`:

```{r list8, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

length(-------)
```

```{r list8-solution}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

length(my_list$cat_names)

#this also works

length(my_list[[1]])

#this also works

length(my_list[['cat_names']])
```

### `purrr::pluck()`

The `purrr` package has an additional way to extract list elements: `purrr::pluck()`. 

```{r pluck, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list %>%
  pluck("cat_names")
```

This can be a handy way to extract something from a list within a list element.

## Putting things into Lists

The slots in a `list` are not fixed. We can take an already made list and add to it.

```{r}

```


## *homogeneous* versus *heterogeneous* lists

When we automate a repetitive process, we usually assume that the lists that we process are *homogeneous*. That is, the data type of the list element is the same for each slot in the list.

However, another common pattern is to return a *heterogenous* list from a function. For example, the `lm()` function actually returns a list.

```{r}

```



## `data.frame`s are list-like

One last note. You might know that you can access columns in a `data.frame` using `$`. This is because a `data.frame` is a special *instance* of a list.

Hence, you can use all of the above accessors to manipulate variables in a `data.frame`.

```{r echo=TRUE}
data(mtcars)

length(mtcars)

mtcars$mpg

mtcars[[1]]

mtcars[["mpg"]]
```

A common pattern in Base-R for filtering that you should be aware of is using the `$` operator for filtering on rows:

```{r echo=TRUE}
mtcars[mtcars$cyl == 8,]

```

## Lists and Vectors

Vectors are a special instance of lists, in that each element can have a name. However, they are known as an *atomic* data type in R. They each have format restrictions.

Let's do something with the three data files in the `data` folder. First, we need to use a command called `list.files()` to list the files in the data folder.

