---
title: "Learning Lists"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(purrr)
library(palmerpenguins)
data(penguins)
dfs <- readRDS("data/list_structure.rds")

get_first_element <- function(input_list){
  return(input_list[[1]])
}

my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))
```

## Learning Objectives

- **Learn** about lists as a general purpose data structure
- **Learn** and **utilize** list properties 
- **Access** list elements using `$` and `[[]]`
- **Understand** the difference between homogeneous and heterogeneous lists
- **Use** `purrr::pluck()` to access list elements
- **Understand** how `data.frames()` are list-like
- **Understand** how to apply a function to a list.

## What are lists?

Lists are a general purpose data structure in R. They are very flexible, in that they have *slots*. A *slot* can contain anything: it could contain a `vector`, a `data.frame`, or even another list!

Unlike vectors, a `list` can also be heterogeneous: each slot can contain a different data type.

We'll start with a making a list and understand different properties about it.

```{r list1, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list
```

`my_list` has some properties and ways to access things in it that are really useful. The first is that `my_list` has a `length()`

```{r list2, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

length(my_list)
```

Additionally, because the elements are named, it also will give you the names:

```{r list3, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

names(my_list)
```

## Accessing Elements of Lists

We can access a single element in the list using a name and `$`:

```{r list4, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list$cat_names
```

You can also access an element using the `[[]]` (double brackets) accessor to access an element in order. Here we're accessing the first element of a list.

```{r list5, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list[[1]]
```

You can also pass a name here as a `character`, if the slots are named.

```{r list6, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list[["hedgehog_names"]]
```

Beware of the difference between `[[]]` and `[]`. The `[]` accessor returns a list of length 1 that is named `hedgehog_names`, not the element in it:

```{r list7, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list["hedgehog_names"]
```

In most cases, you want to use `[[]]`.



### Your Turn

Find the length of the `cat_names` element in `my_list`:

```{r list8, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

length(-------)
```

```{r list8-solution}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

# all of these are acceptable
length(my_list[[1]])
length(my_list$cat_names)
length(my_list[["cat_names"]])

```


### `purrr::pluck()`

The `purrr` package has an additional way to extract list elements: `purrr::pluck()`. This can be a handy way to get a single list element.

```{r pluck, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list %>%
  pluck("cat_names")
```

This can be a handy way to extract something from a list within a list element.

### Your Turn

Try the following code. What did it return? 

Did it return a `data.frame`, or a vector?

```{r pluck2, exercise=TRUE}
penguins %>%
  pluck("species")
```


## Putting things into Lists

The slots in a `list` are not fixed. We can take an already made list and add a new slot to it using `<-`:

```{r adding_list, exercise=TRUE}
my_list <- list(cat_names = c("Morris", "Julia"), 
                hedgehog_names = "Spiny", 
                dog_names = c("Rover", "Spot"))

my_list$raccoon_names <- c("Rocky")

my_list
```


## *homogeneous* versus *heterogeneous* lists

When we automate a repetitive process, we usually assume that the lists that we process are *homogeneous*. That is, the data type of the list element is the same for each slot in the list. In our case, usually the data.type will be a `data.frame`

However, another common pattern is to return a *heterogenous* list from a function. For example, the `lm()` function actually returns a list.

```{r output_sum, exercise=TRUE}
output <- lm(body_mass_g ~ bill_length_mm, data=penguins)
sum_output <- summary(output)

#show that sum_output is a list
names(sum_output)

sum_output$coefficients
```

You often return a heterogeneous list when you are returning multiple objects from a function:

```{r function1, exercise=TRUE}
run_penguin_model <- function(species = "Torgersen") {
  
  subset_penguins <- penguins %>%
    filter(species == species)
    
  output <- summary(lm(body_mass_g ~ bill_length_mm, data=subset_penguins))
  
  out_list <- list(coefficients = output$coefficients, call = output$call)
  
  return(out_list)
}

output_model <- run_penguin_model("Torgersen")

output_model
```


### Your Turn

From `output_model`, return the `call` slot. What is in it?

```{r call, exercise=TRUE}
output_model$-------

```

```{r call-solution}
output_model$call
```


## `data.frame`s are list-like

One last note. You might know that you can access columns in a `data.frame` using `$`. This is because a `data.frame` is a special *instance* of a list.

Hence, you can use all of the above accessors to manipulate variables in a `data.frame`.

```{r df, exercise=TRUE}
data(mtcars)

length(mtcars)

mtcars$mpg

mtcars[[1]]

mtcars[["mpg"]]
```

A common pattern in Base-R for filtering that you should be aware of is using the `$` operator for filtering on rows. The `tidyverse` simplifies this with `filter()`, but this pattern is often used with objects in Bioconductor.

```{r df2, echo=TRUE}
mtcars[mtcars$cyl == 8,]
```

## Applying a function to a list using `map()`

Let's make a function:

```{r get_first, exercise=TRUE}
get_first_element <- function(input_list){
  return(input_list[[1]])
}

```

We can *apply* this function to each element of a list. This is usually easiest if the list is homogeneous.

The function we use is called `map()`. `map()` usually takes two inputs:

- the list to process
- the function to process them with

Note that we don't call `get_first_element()` with the parentheses. 

```{r get_first2, exercise=TRUE}
new_list <- map(my_list, get_first_element)
new_list
```

Next time we will use `map()` to load in multiple `data.frame`s from files in a folder.

## Test Yourself

### Question 1 (2 Points)

Load the following list using `readRDS()` - an RDS file is a way to save a single R object as a file (you can write them using `saveRDS()`). 

What are the names of slots on the list? (show the function)

```{r dfs1, exercise=TRUE}
dfs


```

What is in each slot? Just put your answer below.

```{r dfs1-solution}
dfs
names(dfs)
dfs[[1]]

```



### Question 2 (2 Points)

Use `bind_rows()` on the first and third element of `dfs`:

```{r bind_rows, exercise=TRUE}
bind_df <- bind_rows(------, -------)
nrow(bind_df)
```

```{r bind_rows-solution}
bind_df <- bind_rows(dfs[[1]], dfs[[3]])
nrow(bind_df)

```

### Question 3 (2 Points)

Find the maximum `age_at_diagnosis` for `dfs[["smoke_2.xlsx"]]`: (remember, you can use `summarize`)

```{r max_age, exercise=TRUE}
dfs[["smoke_2.xlsx"]] %>%
  summarize(max_age = --------------)
```

```{r max_age-solution}
dfs[["smoke_2.xlsx"]] %>%
  summarize(max_age = max(age_at_diagnosis))

```

